Master's Project Journal

### 2019-04-11 thesis/Background.tex
more pngs and svgs

The main drivers of performance in \acp{SOC} are the number of transistors on a chip, which is correlated with the number of calculations that can be carried out simultaneously, and the frequency at which the device operates, which determines the number of calculations performed per second.
Moore's Law, based on the famous observation by Gordon Moore in 1965\cite{moore1965cramming}, predicted a doubling in the transistor count of \ac{IC}s per year for the forthcoming decade. This behaviour has carried on to this day as a result of the ever decreasing size of transistors, and has only begun to slow down in recent years.
However, as the number of transistors on a chip has increased roughly following Moore's Law, the increase in clock frequency has not been able to follow a similar linear trajectory, having remained roughly equivalent for the last number of years \cite{ross2008cpu}, indeed the clock speed of the Intel Core family of \ac{CPU}s has not changed since their introduction in 2009 \cite{intelark}.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.4]{../eldar_last_30_yrs}
	\caption{Frequency of the Intel microprocessors over past 30 years \cite{zianbetov2013phd}.}
	\label{fig:eldar_last_30_yrs}
\end{figure}\\
This plateauing of clock frequency has been caused by high power consumption due to the demands placed by the global distribution of a high frequency clock, often the single biggest consumer of power on the chip \cite{tiwari1998reducing}.
With the growth of the \ac{IOT} market where low power devices are desirable, with many of the emerging uses of \ac{SOC}s being portable and thus without a permanent power source, high power consumption goes directly against one of the key pillars of the technology. This forces many of these devices to use lower performance hardware in order to reduce the power consumption, and increase the battery life, of their devices.

@@

\section{The Impact of Clocking Errors}
In Figure \ref{fig:eldar_why_precise_clocking} the data path between two synchronously clocked registers is shown, with the circuit's function being carried out by the combinatorial network between the registers.
Each register has a setup time, which represents the amount of time that the input value to a register must remain constant before the clock edge, and a hold time, the time for which the input must remain constant after a clock edge.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.6]{../eldar_why_precise_clocking}
	\caption{Data Flow in a Clocked System \cite{zianbetov2013phd}.}
	\label{fig:eldar_why_precise_clocking}
\end{figure}

@@

\section{Traditional Solutions}
A number of traditional solutions exist which provide \ac{GSLS} clocking systems, using a variety of techniques. The most simple of these implement clock distribution systems that are symmetrical in order to distribute a centrally generated clock signal to all areas of the chip at the same phase. These systems are named in accordance with their geometry, with the most common variants being branch, X or H trees. 
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.33]{../trees}
	\caption{H and Branch Tree Clock Distribution Systems.}
	\label{fig:trees}
\end{figure}

@@

A mesh clock distribution network is an alternate design where the clock is delivered using a Cartesian grid of distribution lines. Compared to a tree type system, the variation in skew seen with a clock mesh is inversely proportional to the density of the grid while the sources of jitter remain identical. According to Abdelhadi \textit{et al} (2010) clock meshes ``\textit{achieve low and deterministic skew, low skew variations, and low jitter}'', all desirable characteristics for a clock distribution system. However they dissipate more power due to extra capacitive loading, attributable to vast number of lines required to form the grid. Similarly mesh distribution networks suffer from potential mismatch in production and alleviation through increasing of the dimensions of interconnects will, as with a tree type system, lead to higher power consumption \cite{abdelhadi2010timing}. 
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../tex_files/eldar_mesh}
	\caption{Mesh Clock Distribution System \cite{zianbetov2013phd}.}
	\label{fig:mesh}
\end{figure}
Alternative designs replace the electrical lines used in the tree networks with waveguides for optical signals, with only the distribution in the local area carried out using regular wires. This technique presents many advantages \cite{chen2006chip}: optical clock delivery is immune to the noise sources that affect electrical clock distribution systems, consume less power and do not suffer from the electrical losses present in a regular tree system.

@@

One such method is a network of oscillators as in Figure \ref{fig:mizuno1998noise} which uses coupled \ac{PLL} to generate local clocks. Here the output of a leaf node is compared with an external reference and the operating frequency of each \ac{VCO} tuned based on the result.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../mizuno1998noise}
	\caption{Coupled Oscillator Clock Delivery Circuit \cite{mizuno1998noise}.}
	\label{fig:mizuno1998noise}
\end{figure}
The advantage of this method is the simplicity of the feedback network, requiring just the divided clock output from a single leaf node. The \ac{VCO}s then adjusted by the control voltage, $v_c$, which needs delivery to all areas of the chip. However, this is a regular signal and as such does not suffer from skew or jitter. This alleviates the need for a power hungry distribution circuit, while also being more noise-immune than the transmission of a high frequency clock. However this design still suffers from clock variation as all \ac{VCO}s are fed the same control voltage, and thus the manufacturing tolerance issues present in conventional designs persists here also. This is acknowledged by the authors:
\begin{quotation}
	\textit{Unfortunately, as with the conventional ... method, distributing the \ac{VCO}s over the entire chip causes the problem that jitter and skew are increased by variations in the fabrication process (static), temperature, and power supply (dynamic)} \cite{mizuno1998noise}.
\end{quotation}
This type of multi-oscillator design is implemented by analogue circuits, and as a result not only are the clock signals, but also the control signals are liable to variation due to noise, fabrication mismatch and power supply dynamics.

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../gutnik2000active}
	\caption{PLL Network Topology \cite{gutnik2000active}.}
	\label{fig:gutnik2000active}
\end{figure}

@@

In this design the oscillators are once again laid out in a Cartesian grid, with each node coupled to their neighbours in phase. As this is now a digital system the coupling is carried out using digital phase comparators, which attempt to measure the phase difference between two oscillators. Figure \ref{fig:eldar_node} shows high level detail of the architecture of both the entire clocking system and that of an individual node in the design. The digital nature of this architecture brings with it a number of advantages over traditional analogue implementations, as it can benefit from advancements in digital circuit design suites, be reconfigurable and programmable and has a significantly greater immunity to perturbations inherent to its digital nature, as the exact voltage of signals is of no importance \cite{zianbetov2013phd}. This last advantage is of particular use in a digital environment, as otherwise there is potential for clock degradation resulting from switching of transistors. The drawback of the switch to a digital architecture however is the presence of quantisation. Analogue designs both deliver continuous control signals to the oscillators and have a continuous phase detection capability, unlike a digital system where these actions are carried out with fixed resolution.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.4]{../ccirc_2013_arch}
	\caption{Architecture of the ADPLL network and of a single node \cite{zianbetov2013distributed}.}
	\label{fig:eldar_node}
\end{figure}

Looking at the design of a given node it is notable that the function carried out by the Error Combiner is akin to a average, therefore as mentioned by Pratt and Nyugen, there is potential to a mode locked equilibrium in which the oscillators are not synchronised. In their paper, they presented a method where initial start-up was performed uni-directionally and, once all nodes are close to alignment, full connectivity could be restored, however this was not viable in an analogue system as reconfigurability was not an option \cite{pratt1995distributed}. In creating an entirely digital system, Zianbetov and Shan could exploit reconfigurability and implement a uni-directional start-up and thus avoid the problem of convergence into a mode locked state, without having to design a non-linear phase detector.

\section{\acl{ADPLL} Architecture}
As indicated in Figure \ref{fig:mulkeen_pll} the three main building blocks of a conventional \ac{PLL} are the \ac{PD}, \ac{LF} and \ac{VCO}. In an \ac{ADPLL} these blocks are then replaced by their digital counterparts, necessitating quantisation in order to remain physically realisable. The ``All-Digital'' moniker is a misnomer as the oscillator and \acl{PD} are usually both implemented by mixed signal circuits.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{../tex_files/mulkeen_pll}
	\caption{Block Diagram of a \acl{PLL}, \textit{Wireless Systems Notes}, B. Mulkeen (2017).}
	\label{fig:mulkeen_pll}
\end{figure}

\subsection{Digitally Controlled Oscillator}
In a digital system there are a very limited number of voltages representable, most commonly just two, so using a voltage to control the oscillator is not a viable strategy. Instead a fixed bit width signal is used to control the oscillator's period, selecting the number of inverters in a ring oscillator or the varactor configuration of a travelling wave oscillator \cite{chen2011rotary}. The decisions made in the design of the \ac{DCO}, or \ac{NCO}, determine many of the other \ac{ADPLL} parameters. While tuning range and centre frequency, as well as linearity, carry over from the analogue counterpart, a \ac{DCO} also has a frequency step which in combination with the bit width of the control signal determines the range over which the oscillator can be tuned. Figure \ref{fig:my_ring} illustrates a basic ring oscillator design. A ring oscillator is an inherently unstable circuit composed of an odd number of inverters connected in a circle, which allows a signal to propagate infinitely, with the signal at any point in the circuit appearing as a square wave. The half-period of this oscillator is the time taken for the signal to propagate once through the chain, $n$ times the propagation delay through one inverter. The frequency of operation can then be set by modulating the length of the chain, in steps of two inverters to maintain an odd number, by means of \texttt{f\_select}. The main impact of output frequency quantisation is that only frequencies which are integer multiples of the frequency step away from the centre frequency can be easily reproduced, with intermediate values only obtainable in a manner akin to Fractional-N synthesis with the control code toggling back and forth. This acts as a source of jitter in the system.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.275]{../inverter_chain}
	\caption{Basic Ring/Inverter Chain Oscillator.}
	\label{fig:my_ring}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.35]{../shan_bb_pd}
	\caption{Bang-bang phase/frequency detector architecture. \cite{shan2014phd}.}
	\label{fig:shan_bb_pd}
\end{figure}

\subsection{Digital Loop Filter}
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.45]{../pi_simple}
	\caption{Basic PI Controller Architecture.}
	\label{fig:my_simple_pi}
\end{figure}%TODO ask mulkeen if this is IIR
The \acl{LF} in an \ac{ADPLL} can be implemented as a \ac{PI} controller, as only a low-pass filter is required, such as that in Figure \ref{fig:my_simple_pi}. In the case of a node in an \ac{ADPLL} network the input of this filter is a weighted of the phase difference relative to the neighbouring local clocking areas. In one example topology a digital system the proportional section can be implemented by a simple multiplier, whereas the integral path is constructed by adding the result of a multiplication by the proportional gain to an accumulator. This delayed summation can be easily implemented by an accumulator to which the current value of the multiplication is added each cycle, and as such the system has an infinite impulse response. The value of these gains determine the response and stability of the \ac{ADPLL} network. The transfer function of such a controller is given by \cite{shan2014phd}:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}
\end{equation*}
It has been found by Koskin \textit{et al} (2018) that stable operation can be achieved when the integral gain, $k_i$, is less than half the proportional gain, $k_p$ \cite{koskin2018generation}. In the same study a range of values was found which would produce low jitter operation of the network. As these values are all less than one, the filter must implement fixed point arithmetic in an effort to maintain the simplicity of the clock distribution network, rather than incurring the complexity penalty of floating point calculations.

>>>>

The main drivers of performance in \acp{SOC} are the number of transistors on a chip, which is correlated with the number of calculations that can be carried out simultaneously, and the frequency at which the device operates, which determines the number of calculations performed per second.
Moore's Law, based on the famous observation by Gordon Moore in 1965\cite{moore1965cramming}, predicted a doubling in the transistor count of \ac{IC}s per year for the forthcoming decade. This behaviour has carried on to this day as a result of the ever decreasing size of transistors, and has only begun to slow down in recent years.
However, as the number of transistors on a chip has increased roughly following Moore's Law, the increase in clock frequency has not been able to follow a similar linear trajectory, having remained roughly equivalent for the last number of years \cite{ross2008cpu}, indeed the clock speed of the Intel Core family of \ac{CPU}s has not changed since their introduction in 2009 \cite{intelark}.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.4]{../eldar_last_30_yrs}
	\caption[Frequency of the Intel microprocessors over past 30 years]{Frequency of the Intel microprocessors over past 30 years \cite{zianbetov2013phd}.}
	\label{fig:eldar_last_30_yrs}
\end{figure}\\
This plateauing of clock frequency has been caused by high power consumption due to the demands placed by the global distribution of a high frequency clock, often the single biggest consumer of power on the chip \cite{tiwari1998reducing}.
With the growth of the \ac{IOT} market where low power devices are desirable, with many of the emerging uses of \ac{SOC}s being portable and thus without a permanent power source, high power consumption goes directly against one of the key pillars of the technology. This forces many of these devices to use lower performance hardware in order to reduce the power consumption, and increase the battery life, of their devices.

@@

\section{The Impact of Clocking Errors}
In Figure \ref{fig:eldar_why_precise_clocking} the data path between two synchronously clocked registers is shown, with the circuit's function being carried out by the combinatorial network between the registers.
Each register has a setup time, which represents the amount of time that the input value to a register must remain constant before the clock edge, and a hold time, the time for which the input must remain constant after a clock edge.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.6]{../eldar_why_precise_clocking}
	\caption[Data flow in a clocked system]{Data flow in a clocked system \cite{zianbetov2013phd}.}
	\label{fig:eldar_why_precise_clocking}
\end{figure}

@@

\section{Traditional Solutions}
A number of traditional solutions exist which provide \ac{GSLS} clocking systems, using a variety of techniques. The most simple of these implement clock distribution systems that are symmetrical in order to distribute a centrally generated clock signal to all areas of the chip at the same phase. These systems are named in accordance with their geometry, with the most common variants being branch, X or H trees. 
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.33]{../trees}
	\caption[H and branch tree clock distribution systems]{H and branch tree clock distribution systems.}
	\label{fig:trees}
\end{figure}

@@

A mesh clock distribution network is an alternate design where the clock is delivered using a Cartesian grid of distribution lines. Compared to a tree type system, the variation in skew seen with a clock mesh is inversely proportional to the density of the grid while the sources of jitter remain identical. According to Abdelhadi \textit{et al} (2010) clock meshes ``\textit{achieve low and deterministic skew, low skew variations, and low jitter}'', all desirable characteristics for a clock distribution system. However they dissipate more power due to extra capacitive loading, attributable to vast number of lines required to form the grid. Similarly mesh distribution networks suffer from potential mismatch in production and alleviation through increasing of the dimensions of interconnects will, as with a tree type system, lead to higher power consumption \cite{abdelhadi2010timing}. 
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../tex_files/eldar_mesh}
	\caption[Mesh clock distribution system ]{Mesh clock distribution system \cite{zianbetov2013phd}.}
	\label{fig:mesh}
\end{figure}
Alternative designs replace the electrical lines used in the tree networks with waveguides for optical signals, with only the distribution in the local area carried out using regular wires. This technique presents many advantages \cite{chen2006chip}: optical clock delivery is immune to the noise sources that affect electrical clock distribution systems, consume less power and do not suffer from the electrical losses present in a regular tree system.

@@

One such method is a network of oscillators as in Figure \ref{fig:mizuno1998noise} which uses coupled \ac{PLL} to generate local clocks. Here the output of a leaf node is compared with an external reference and the operating frequency of each \ac{VCO} tuned based on the result.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../mizuno1998noise}
	\caption[Coupled oscillator clock delivery circuit]{Coupled oscillator clock delivery circuit \cite{mizuno1998noise}.}
	\label{fig:mizuno1998noise}
\end{figure}
The advantage of this method is the simplicity of the feedback network, requiring just the divided clock output from a single leaf node. The \ac{VCO}s then adjusted by the control voltage, $v_c$, which needs delivery to all areas of the chip. However, this is a regular signal and as such does not suffer from skew or jitter. This alleviates the need for a power hungry distribution circuit, while also being more noise-immune than the transmission of a high frequency clock. However this design still suffers from clock variation as all \ac{VCO}s are fed the same control voltage, and thus the manufacturing tolerance issues present in conventional designs persists here also. This is acknowledged by the authors:
\begin{quotation}
	\textit{Unfortunately, as with the conventional ... method, distributing the \ac{VCO}s over the entire chip causes the problem that jitter and skew are increased by variations in the fabrication process (static), temperature, and power supply (dynamic)} \cite{mizuno1998noise}.
\end{quotation}
This type of multi-oscillator design is implemented by analogue circuits, and as a result not only are the clock signals, but also the control signals are liable to variation due to noise, fabrication mismatch and power supply dynamics.

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.7]{../gutnik2000active}
	\caption[PLL network topology ]{PLL network topology \cite{gutnik2000active}.}
	\label{fig:gutnik2000active}
\end{figure}

@@

In this design the oscillators are once again laid out in a Cartesian grid, with each node coupled to their neighbours in phase. As this is now a digital system the coupling is carried out using digital phase comparators, which attempt to measure the phase difference between two oscillators. Figure \ref{fig:eldar_node} shows high level detail of the architecture of both the entire clocking system and that of an individual node in the design. The digital nature of this architecture brings with it a number of advantages over traditional analogue implementations, as it can benefit from advancements in digital circuit design suites, be reconfigurable and programmable and has a significantly greater immunity to perturbations inherent to its digital nature, as the exact voltage of signals is of no importance \cite{zianbetov2013phd}. This last advantage is of particular use in a digital environment, as otherwise there is potential for clock degradation resulting from switching of transistors. The drawback of the switch to a digital architecture however is the presence of quantisation. Analogue designs both deliver continuous control signals to the oscillators and have a continuous phase detection capability, unlike a digital system where these actions are carried out with fixed resolution.
\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.4]{../ccirc_2013_arch}
	\caption[Architecture of the ADPLL network and of a single node]{Architecture of the ADPLL network and of a single node \cite{zianbetov2013distributed}.}
	\label{fig:eldar_node}
\end{figure}

Looking at the design of a given node it is notable that the function carried out by the Error Combiner is akin to a average, therefore as mentioned by Pratt and Nyugen, there is potential to a mode locked equilibrium in which the oscillators are not synchronised. In their paper, they presented a method where initial start-up was performed uni-directionally and, once all nodes are close to alignment, full connectivity could be restored, however this was not viable in an analogue system as reconfigurability was not an option \cite{pratt1995distributed}. Javidan \textit{et al} (2011) found that this was in fact the case, and stated:
\begin{quote}
	[Mode-locking] is solved in a simple and original way, by a dynamic reconfiguration of the network interconnection topology at the starting stage.
\end{quote}
In creating an entirely digital system, Zianbetov and Shan could easily reconfigure the network topology and by implement a uni-directional start-up avoid the problem of convergence into a mode locked state, without having to design a non-linear phase detector.

\section{\acl{ADPLL} Architecture}
As indicated in Figure \ref{fig:mulkeen_pll} the three main building blocks of a conventional \ac{PLL} are the \ac{PD}, \ac{LF} and \ac{VCO}. In an \ac{ADPLL} these blocks are then replaced by their digital counterparts, necessitating quantisation in order to remain physically realisable. The ``All-Digital'' moniker is a misnomer as the oscillator and \acl{PD} are usually both implemented by mixed signal circuits.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{../tex_files/mulkeen_pll}
	\caption[Block fiagram of a \acl{PLL}]{Block fiagram of a \acl{PLL}, \textit{Wireless Systems Notes}, B. Mulkeen (2017).}
	\label{fig:mulkeen_pll}
\end{figure}

\subsection{Digitally Controlled Oscillator}
In a digital system there are a very limited number of voltages representable, most commonly just two, so using a voltage to control the oscillator is not a viable strategy. Instead a fixed bit width signal is used to control the oscillator's period, selecting the number of inverters in a ring oscillator or the varactor configuration of a travelling wave oscillator \cite{chen2011rotary}. The decisions made in the design of the \ac{DCO}, or \ac{NCO}, determine many of the other \ac{ADPLL} parameters. While tuning range and centre frequency, as well as linearity, carry over from the analogue counterpart, a \ac{DCO} also has a frequency step which in combination with the bit width of the control signal determines the range over which the oscillator can be tuned. Figure \ref{fig:my_ring} illustrates a basic ring oscillator design. A ring oscillator is an inherently unstable circuit composed of an odd number of inverters connected in a circle, which allows a signal to propagate infinitely, with the signal at any point in the circuit appearing as a square wave. The half-period of this oscillator is the time taken for the signal to propagate once through the chain, $n$ times the propagation delay through one inverter. The frequency of operation can then be set by modulating the length of the chain, in steps of two inverters to maintain an odd number, by means of \texttt{f\_select}. The main impact of output frequency quantisation is that only frequencies which are integer multiples of the frequency step away from the centre frequency can be easily reproduced, with intermediate values only obtainable in a manner akin to Fractional-N synthesis with the control code toggling back and forth. This acts as a source of jitter in the system.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.275]{../inverter_chain}
	\caption[Basic Ring/Inverter Chain Oscillator]{Basic ring/inverter chain oscillator.}
	\label{fig:my_ring}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.35]{../shan_bb_pd}
	\caption[Bang-bang phase/frequency detector architecture]{Bang-bang phase/frequency detector architecture \cite{shan2014phd}.}
	\label{fig:shan_bb_pd}
\end{figure}

\subsection{Digital Loop Filter}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../simple_pi}
	\caption[Basic \ac{PI} controller architecture]{Basic \ac{PI} controller architecture.}
	\label{fig:my_simple_pi}
\end{figure}%TODO ask mulkeen if this is IIR
The \acl{LF} in an \ac{ADPLL} can be implemented as a \ac{PI} controller, as only a low-pass filter is required, such as that in Figure \ref{fig:my_simple_pi}. In the case of a node in an \ac{ADPLL} network the input of this filter is a weighted of the phase difference relative to the neighbouring local clocking areas. In one example topology a digital system the proportional section can be implemented by a simple multiplier, whereas the integral path is constructed by adding the result of a multiplication by the proportional gain to an accumulator. This delayed summation can be easily implemented by an accumulator to which the current value of the multiplication is added each cycle, and as such the system has an infinite impulse response. The value of these gains determine the response and stability of the \ac{ADPLL} network. The transfer function of such a controller is given by \cite{shan2014phd}:
\begin{equation*}
	H(z) = \alpha + \beta\frac{1}{1-z^{-1}} = \frac{(\alpha + \beta) - \alpha z^{-1}}{1-z^{-1}}
\end{equation*}
It has been found by Koskin \textit{et al} (2018) that stable operation can be achieved when the integral gain, $k_i$, is less than half the proportional gain, $k_p$ \cite{koskin2018generation}. In the same study a range of values was found which would produce low jitter operation of the network. As these values are all less than one, the filter must implement fixed point arithmetic in an effort to maintain the simplicity of the clock distribution network, rather than incurring the complexity penalty of floating point calculations.
